---
title: "Practical Fall 2019"
author: "Sunyimeng Lei"
date: "10/28/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Documents/UMich_Bio201_F19/")
```

As with each lab, add "completed" to the Rmd filename; update the name in the author line above; update the location of the working directory as necessary. 

# Load packages
```{r Load packages, eval=TRUE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

# Part 1 - concepts 
Type the answers to questions 1-5 in the space below. 

### Question 1

Given vector v1 below, will it be read as character or numeric? Why does this happen? [1 point]
```{r}
v1 <- c("one", "2", "three", "4")
```

<type answer here> 

character. It's a vector and all elements need to hav the same mode and numerics should comprimise themselves to character if ther's characters in the vector.

### Question 2

dataframe1 has 323 rows and dataframe2 has 345 rows. After an inner join the resulting data fame has 312 rows. Why does the result have fewer rows? [1 point]

<type answer here> 
There's some values in dataframe1 do not match what's in dataframe2 and vice versa.

### Question 3

What are the three assumptions for a t-test? Explain what to do if the variance assumption is violated. Explain what to do if the normality assumption is violated. [3 points]

<type answer here> 
* Relatively large sample size (usually > 30)
* The samples have to come from a normal distribution
* We are also going to check if the variances of the groups are equal, because this will determine some of the t.test() arguments

If variance assumption is violated, in t-test, change the argument of var.equal to FALSE. If normality assumption is violated, first check how much the distrubtion deviates from normal with a histogram. If not normal,then use the non-parametric test.(two-sample wilcoxon test)

### Question 4

You have a data frame (called butyrate_df) with the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Avg_butyrate". If this data frame is piped to each of the functions below, evaluate if the description of the results is true or false. [3 points]
```{r eval=FALSE, include=FALSE}
butyrate_df %>%
  filter(Semester == "Fall2019")
# Result1 = keeps only data from Fall 2019 semester

butyrate_df %>%
  filter(Avg_butyrate <= 47.0)
# Result2 = keeps only values equal to 47 mmol/kg 

butyrate_df %>%
  group_by(Semester) %>%
  summarize(New_column = median(Avg_butyrate))
# Result3 = new_column contains the median of each participant 
```

<type answer here> 
Result1:True
Result2:False
Result3:False

### Question 5

something about NAs

The data frame (called breath_df) has the columns: "Participant_ID", "Study_week", "Semester", "Supplement_consumed", "Methane", "Hydrogen", "Carbon_dioxide". This data frame is used for the code below, explain what will happen to any NAs in each of the shown calculations, and how the NAs will effect the underlying data. [2 points]
```{r eval=FALSE, include=FALSE}
# Option 1
breath_df %>%
  filter(!is.na(Methane)) %>% 
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane))

# Option 2
breath_df %>%
  group_by(Participant_ID, Study_week) %>%
  summarize(Avg_methane = mean(Methane, na.rm = TRUE))
```

<type answer here> 

The row containing NA in the column "Methane" will be removed the average_methane will be the mean of each participant per week. Any NAs in ang groups of data will return to a NA in the output.


# Part 2 - tidy data 

### Question 6

Find the data frame matching your unique name in the GitHub repository. Find the breath gas and taxa abundance data frames. Download and import all into RStudio and use these data frames to complete the rest of the practical. Name the dataframe to match your uniqname, name the breath gas data frame "breath", name the taxa abundance data frame "taxa". [1 point]
```{r}
# data import code 
lsym <- read_delim(file = "Practial/Raw_data/lsym.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA"))
breath <- read_delim(file = "Practial/Raw_data/breath.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA"))
taxa <- read_delim(file = "Practial/Raw_data/taxa.txt",
                        delim = "\t", col_names = TRUE, trim_ws = TRUE, 
                        na = c("", "NA"))

```

Update column names in all files match the course conventions. [1 point]
```{r}

 lsym<-rename_all(lsym,tolower)
taxa<-rename_all(taxa,tolower) %>% 
rename(relative_abundance = fract)

breath<-rename_all(breath,tolower)

```

View the supplement types in your semester. Are they all entered in the same style? If not update them to match. [1 point]
```{r}

  
lsym[["supplement_consumed"]] <- recode(.x=lsym[["supplement_consumed"]], "transition HiMaize"="transition_HiMaize")

count(lsym, supplement_consumed)


```

### Question 7

What are the dimensions of each data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(lsym) #482rows, 9columns
dim(taxa) #46681rows, 6columns
dim(breath) #2668rows, 5columns

```

Select a join that will drop as few data points as possible. Join the data frame containing the participant data with the breath gas data. Name this "joined_df". [1 point]
```{r}
joined_df<-full_join(lsym,breath,by=c("participant_id","study_week","semester"))

```

What are the dimensions of the joined data frame? Write the results as a comment after the function. [1 point]
```{r}
dim(joined_df) #3159rows, 11columns
```

### Question 8

Calculate the total SCFAs. [1 point]
```{r}
joined_df %>% 
  mutate(scfa_total_mmol_kg=butyrate_mmol_kg+acetate_mmol_kg+propionate_mmol_kg)

```

Calculate the weekly mean of each SCFA of each participant. Calculate the weekly mean methane and hydrogen levels of each participant. [1 point]
```{r}
wkly_mean<-joined_df %>%
  filter(semester=="Winter2018") %>% 
  mutate(scfa_total_mmol_kg=butyrate_mmol_kg+acetate_mmol_kg+propionate_mmol_kg) %>% 
  group_by(participant_id,study_week,supplement_consumed) %>% 
  summarize(but_wkly = mean(butyrate_mmol_kg, na.rm = TRUE),
            ace_wkly=mean(acetate_mmol_kg,na.rm=TRUE),
pro_wkly=mean(propionate_mmol_kg,na.rm = TRUE),
methane=mean(ch4, na.rm=TRUE),
hydrogen=mean(h2,na.rm = TRUE))
  wkly_mean
```

What are the dimensions of the data frame with the avearges? Write the results as a comment after the function.
```{r}
dim(wkly_mean)
# 183rows   8columns
```

Join this data frame with the one called "taxa". Name the new data frame "all_data". What are the dimensions of this new data frame? [1 point]
```{r}
all_data<-inner_join(wkly_mean,taxa,by=c("study_week","participant_id"))
dim(all_data)
#4132    12

#according to the question guide, we cannot have the data of all participants with regard to breath hydrogen and methane for the use of question 10. In question 10 i full join the taxa and joined_df to make a whole data set we need.
```

Save all data frames generated to your GitHub page. [1 point]
```{r}
write_delim(all_data, path = "Practial/curated_data/all_data.txt", delim = "\t")
write_delim(joined_df, path = "Practial/curated_data/joined_df.txt", delim = "\t")
write_delim(wkly_mean, path = "Practial/curated_data/wkly_mean.txt", delim = "\t")
```

# Part 3 - statistics & graphics 

Update all axes labels and figure titles so they are informative and easy to read. Upload final plots to GitHub. When specified, conduct the appropriate tests, and write the statistic (p, F, R value) as a comment. 

### Question 9

If you have more than one type of supplement consumed in your semester, filter for one type to answer Question 9. Plot each SCFA with the study week (weeks 1 and 3 only) on the x-axis. Make each study week a different color. Hint: you may have to use facets, or create individual plots and combine them into a multi-panelled figure. [3 points]
```{r}
plot_but<-wkly_mean %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed=="BRMPS") %>% 
    ggplot(aes(x = study_week, 
             y = but_wkly,
             color = study_week)) + 
  geom_violin() + geom_jitter()+  
  xlab(NULL) + 
  ylab("Butyrate (mmol/kg)") + 
  theme(legend.position = "none")
   
    
    plot_ace<-wkly_mean %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed=="BRMPS") %>% 
    ggplot(aes(x = study_week, 
             y = ace_wkly,
             color = study_week)) + 
  geom_violin() + geom_jitter()+  
  xlab(NULL) + 
  ylab("Acetate (mmol/kg)") + 
  theme(legend.position = "none")
   
    
    plot_pro<-wkly_mean %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed=="BRMPS") %>% 
    ggplot(aes(x = study_week, 
             y = pro_wkly,
             color = study_week)) + 
  geom_violin() + geom_jitter()+  
  xlab(NULL) + 
  ylab("Propionate (mmol/kg)") + 
  theme(legend.position = "none")
   
    
   plot_combined<- plot_grid(plot_ace, plot_but, plot_pro, 
          labels = c("A", "B", "C"),
          nrow = 2, ncol = 2)  
    save_plot(filename = "Practial/figures/plot_combined.pdf",
          plot = plot_combined)

    plot_combined

```

Based on the data presented in the plot you generated above, identify which statistical test you would need to determine if there was a change in SCFA concentration between weeks 1 and 3 of the study. Check the assumptions appropriate for the test. [3 points]
```{r}
# assumptions


wkly_mean %>%
  filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") %>% 
  group_by(study_week) %>% 
  summarise(sample_size = n()) 


but_df<-wkly_mean %>% 
   filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") 
  shapiro.test(but_df$but_wkly) 
ggplot(but_df, aes(x=but_wkly)) +
  geom_histogram() 
qqnorm(but_df$but_wkly); qqline(but_df$but_wkly)

but_wk1<- but_df%>%
  filter(study_week == "week1") 
but_wk3 <- but_df %>%
  filter(study_week == "week3") 
var.test(x = but_wk1$but_wkly, 
         y = but_wk3$but_wkly, 
         alternative = "two.sided")
# Normal, unequal variances for butyrate

ace_df<-wkly_mean %>% 
   filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") 
shapiro.test(ace_df$ace_wkly) 
ggplot(ace_df, aes(x=ace_wkly)) +
  geom_histogram() 
qqnorm(ace_df$ace_wkly); qqline(ace_df$ace_wkly)

ace_wk1<- ace_df%>%
  filter(study_week == "week1") 
ace_wk3 <- ace_df %>%
  filter(study_week == "week3") 
var.test(x = ace_wk1$ace_wkly, 
         y = ace_wk3$ace_wkly, 
         alternative = "two.sided")
# non-normal,equal variance for acetate 

pro_df<-wkly_mean %>% 
   filter(study_week == "week1" | study_week == "week3",
         supplement_consumed == "BRMPS") 
shapiro.test(pro_df$pro_wkly) 
ggplot(pro_df, aes(x=pro_wkly)) +
  geom_histogram() 
qqnorm(pro_df$pro_wkly); qqline(pro_df$pro_wkly)

pro_wk1<- pro_df%>%
  filter(study_week == "week1") 
pro_wk3 <- pro_df %>%
  filter(study_week == "week3") 
var.test(x = pro_wk1$pro_wkly, 
         y = pro_wk3$pro_wkly, 
         alternative = "two.sided")
# non-normal,equal variance for propionate



# test 

t.test(x = but_wk1$but_wkly, 
       y = but_wk3$but_wkly,
       alternative = "less", paired = FALSE, var.equal = FALSE)

wilcox.test(x = ace_wk1$ace_wkly, 
         y = ace_wk3$ace_wkly, 
            alternative = "less", paired = FALSE)
wilcox.test(x = pro_wk1$pro_wkly, 
         y = pro_wk3$pro_wkly, 
            alternative = "less", paired = FALSE)

##Cannot do paired test since participant U541 only has the data for week3.Although all others are mateched



# result?
# but p-value = 0.4672
# ace p-value = 0.3145
# pro p-value = 0.894

## all have large p value. No significant difference in all SCFAs
```

Based on the results of the test, conclude if there was a difference in SCFAs between weeks. 

<type answer here> 
all have large p value. No significant difference(increase) in all SCFAs

### Question 10

Is there a correlation between the abundace of hydrogen generators and breath hydrogen? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
hydrogen_relation<-full_join(joined_df,taxa,by=c("study_week","participant_id")) %>% 
  filter(category=="hydrogen_gen") %>% 
  group_by(participant_id) %>% 
  summarize(hydrogen_gen_abund=sum(relative_abundance),h2_conc=sum(h2)) %>% 
filter(!is.na(h2_conc)) 

hydrogen_relation_plot<-hydrogen_relation %>% 
   ggplot(aes(x = hydrogen_gen_abund,
             y = h2_conc)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("hydrogen generator abundance") + 
  ylab("breath hydrogen") 
hydrogen_relation_plot

save_plot(filename = "Practial/figures/hydrogen_relation.pdf",
          plot = hydrogen_relation_plot)
# test
a10 <- hydrogen_relation %>%
  lm(hydrogen_gen_abund ~ h2_conc, data = .)

summary(a10) 

# result?
#p-value: 3.134e-06
#Multiple R-squared:  0.06327

```

What can you conclude from these results?

They are not linear correlated, although the p-value is small which means significant,which may because we have large sample size.
<type answer here> 

Is there a correlation between the abundace of hydrogen consumers and breath methane? Create a plot that shows the relationship and conduct the appropriate statistical test(s). [2 points]
```{r}
# plot
methane_relation<-full_join(joined_df,taxa,by=c("study_week","participant_id")) %>% 
  filter(category=="hydrogen_consumers") %>% 
  group_by(participant_id) %>% 
  summarize(hydrogen_consumer_abund=sum(relative_abundance),ch4_conc=sum(ch4)) %>% 
filter(!is.na(ch4_conc)) 

methane_relation_plot<-methane_relation %>% 
   ggplot(aes(x = hydrogen_consumer_abund,
             y = ch4_conc)) + 
  geom_point() + 
  geom_smooth(method = "lm", 
              se = FALSE) + 
  xlab("hydrogen consumer abundance") + 
  ylab("breath methane") 
methane_relation_plot

save_plot(filename = "Practial/figures/methane_relation.pdf",
          plot = methane_relation_plot)
# test
b10 <- methane_relation %>%
  lm(hydrogen_consumer_abund ~ ch4_conc, data = .)

summary(b10) 



# result?
#p-value: 2.2e-16
#Multiple R-squared:  0.4374
```

What can you conclude from these results?

<type answer here> 
We have significance since the p value is small, but R-squared does not show good linear fit.(p value is small probably because of large sample size)
### Extra credit:

Is the abundace of hydrogen producers and consumers related? Create a plot, check assumptions, conduct the appropriate statistical test(s), and interpret the results. [1 point]
```{r}
# plot

pro_con<-all_data %>% 
  filter(category=="hydrogen_consumers"|category=="hydrogen_gen") %>% 
 spread(key=category, value=relative_abundance) 

pro_con_plot<-pro_con %>% 
ggplot(aes(x = hydrogen_consumers,
             y = hydrogen_gen)) + 
  geom_point() + 

  xlab("hydrogen consumer abundance") + 
  ylab("hydrogen producer abundace") 
pro_con_plot

save_plot(filename = "Practial/figures/pro_con.pdf",
          plot = pro_con_plot)


# test(s)
e10 <- pro_con %>%
  lm(hydrogen_consumers ~ hydrogen_gen, data = .)

summary(e10) 
```

<type answer here> 


-----
end